---
title: "BI PROJECT TRANSCRIPTOME E1A effects"
author: "EP"
output: html_document
date: "2026-05-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
 
library(data.table)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(ggrepel)
library(dplyr)
library(tximport)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)  
library(enrichplot)     
require(DOSE)
library(tftargets)
library(tidyverse)
```

```{r}
# List Salmon output
samples <- list.files("salmon_quant_24/Salmon quant", full.names = TRUE)
files <- file.path(samples, "quant.sf")
names(files) <- basename(samples)
```


Imort files and count genes:
```{r}
tx2gene <- read.csv("data/tx2gene_symbols.csv")

txi_gene <- tximport(files, type = "salmon", tx2gene = tx2gene, countsFromAbundance="no")
txi_gene_TPM <-  tximport(files, type = "salmon", tx2gene = tx2gene, countsFromAbundance = "lengthScaledTPM")

txi_counts <- data.frame(txi_gene$counts)
txi_TPM <- data.frame(txi_gene_TPM$counts)
write.table(txi_counts, file = "raw_counts_all_samples.txt", sep = "\t", row.names = TRUE, quote = FALSE)
write.csv2(txi_counts, file = "raw_counts_all_samples.csv", row.names = TRUE, quote = FALSE)
write.csv2(txi_TPM, file = "TPM_all_samples.csv", row.names = TRUE, quote = FALSE)
```


Add information about samples
```{r metadata, echo=FALSE}
metadata <- data.frame(
    full_condition = factor(rep(c("ctrl", "E1A", "oxpl", "oxpl_E1A"), each=3,times=2)),
    condition = factor(rep(c("ctrl", "oxpl"), each=6,times=2)),
    e1a_status = factor(rep(c("negative", "positive"), each=3,times=4)),
    cell_line = factor(rep(c("HCT116", "HCT116_oxpl_R"), each=12)),
    stat = factor(rep(c("ctrl", "ctrl_E1A", "oxpl", "oxpl_E1A", "R_ctrl", "R_ctrl_E1A", "R_oxpl", "R_oxpl_E1A"),each=3)))

rownames(metadata) <- colnames(txi_gene$counts)
```

General model for all factors and their interactions
```{r deseq, echo=FALSE}
dds <- DESeqDataSetFromTximport(
    txi = txi_gene,
    colData = metadata,
    design = ~ e1a_status * condition * cell_line # с учетом всех взаимодействий
)

dds <- dds[rowSums(counts(dds)) > 0, ]
dds <- DESeq(dds) 

results <- results(dds, alpha=0.05)
resultsNames(dds)

resOrdered <- results[order(results$pvalue),]
summary(results)

results_df <- as.data.frame(results)

res_clean <- results[!is.na(results$padj) & !is.na(results$log2FoldChange), ]

res_sig <- res_clean[abs(res_clean$log2FoldChange) > 1.5 & res_clean$padj < 0.05, ]
res_sig_df <- as.data.frame(res_sig)
```


```{r deseq_stats, echo=FALSE}
# Get all the contrast names
contrast_names <- resultsNames(dds)
contrast_names <- contrast_names[contrast_names != "Intercept"]  # убираем Intercept

# VST normalization
vsd <- vst(dds, blind = FALSE)

sample_names <- as.character(colData(dds)$stat)

plot_contrast_heatmap <- function(contrast_name, n_top = 30) {
  res_contrast <- results(dds, name = contrast_name)
  
  res_contrast <- res_contrast[!is.na(res_contrast$padj), ]
  
  res_sorted <- res_contrast[order(res_contrast$padj), ]
  top_genes <- head(rownames(res_sorted), n_top)

  top_data <- assay(vsd)[top_genes, ]
  colnames(top_data) <- sample_names
  
  title <- paste("Top", n_top, "genes for contrast:", contrast_name)
  
  pheatmap(top_data,
           scale = "row",
           main = title,
           fontsize_row = 6,
           fontsize_col = 8,
           angle_col = 45,
           treeheight_row = 50,
           treeheight_col = 50,
           border_color = NA,
           silent = FALSE)
}

for (contrast in contrast_names) {
  plot_contrast_heatmap(contrast, n_top = 30)
}
```

```{r}
# e1a_status_positive_vs_negative
res_E1A <- results(dds, name = "e1a_status_positive_vs_negative")

summary(res_E1A)

res_E1A_df <- as.data.frame(res_E1A)
head(res_E1A_df)
```

```{r}
# Number of significant genes
sig_genes <- res_E1A[!is.na(res_E1A$padj) & res_E1A$padj < 0.05, ]
total_sig <- nrow(sig_genes)

# Up / Down
up_genes <- sig_genes[sig_genes$log2FoldChange > 0, ]
down_genes <- sig_genes[sig_genes$log2FoldChange < 0, ]

# padj < 0.05 И |log2FC| > 1
sig_genes_fc1 <- res_E1A[!is.na(res_E1A$padj) & 
                         res_E1A$padj < 0.05 & 
                         abs(res_E1A$log2FoldChange) > 1, ]

up_fc1 <- sig_genes_fc1[sig_genes_fc1$log2FoldChange > 1, ]
down_fc1 <- sig_genes_fc1[sig_genes_fc1$log2FoldChange < -1, ]

cat("Total significant genes:", nrow(sig_genes_fc1), "\n")
cat("  - Up (log2FC > 1):", nrow(up_fc1), "\n")
cat("  - Down (log2FC < -1):", nrow(down_fc1), "\n")
```

At the threshold of ∣log2FC∣>1, we see 117 significant genes, of which 109 are upregulated and only 8 are downregulated. This is consistent with the fact that E1A causes a powerful transcriptional rearrangement of the cell and, in a number of systems, primarily acts as an activator of expression, although it can simultaneously repress individual pathways through interactions with pRb/E2F and p300/CBP.


```{r}
top_genes <- head(order(res_E1A$padj), 30)

top_data <- assay(vsd)[top_genes, ]
colnames(top_data) <- as.character(colData(dds)$stat)

pheatmap(top_data,
           scale = "row",
           main = "Top 30 DEGs: E1A effect",
           fontsize_row = 6,
           fontsize_col = 8,
           angle_col = 45,
           treeheight_row = 50,
           treeheight_col = 50,
           border_color = NA,
           silent = FALSE)
```

On the heatmap, E1A-positive and E1A-negative samples form separate clusters, meaning that the E1A effect affects not just one gene, but the entire expression program.

```{r}
EnhancedVolcano(res_E1A,
    lab = rownames(res_E1A),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'E1A effect (positive vs negative)',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.0,
    labSize = 4.0)
```

In the volcano plot, most significant genes are shifted to the right, indicating increased expression in the E1A-positive group.
Among the most prominent upregulated genes are JUNB, ARC, CIRBP-AS1, SP110, ZFP36, and HSPA1A, while among the downregulated genes, for example, PDXDC2P

This profile is consistent with the fact that E1A is known as a protein that activates AP-1-dependent and other early transcriptional programs, including through interaction with p300/CBP

```{r}
# Top 10 genes by importance
top10_genes <- head(rownames(res_E1A)[order(res_E1A$padj)], 10)

for (gene in top10_genes) {
  ge <- plotCounts(dds, gene, intgroup = "stat", returnData = TRUE)
  
  p <- ggplot(ge, aes(x = stat, y = count, fill = stat)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.8) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = paste("Expression of", gene),
         x = "", y = "Normalized Counts") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 14, face = "bold"),
          plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  
  print(p)
}
```

E1A strongly induces an immediate-early transcriptional program in HCT116 cells, centered on AP-1 and early response genes such as JUNB and EGR1, together with stress/adaptation and chromatin-regulatory factors like DUSP1, HSPA1B, BRD2, and CCN2.


```{r}
# gene_stat
gene_stat <- res_E1A$stat
names(gene_stat) <- rownames(res_E1A)
gene_stat <- na.omit(gene_stat)
gene_stat <- sort(gene_stat, decreasing = TRUE)

# SYMBOL → ENTREZ
mapping <- mapIds(org.Hs.eg.db,
                  keys = names(gene_stat),
                  column = "ENTREZID",
                  keytype = "SYMBOL",
                  multiVals = "first")

gene_stat_clean <- gene_stat[!is.na(mapping)]

# GSEA
gse_GO <- gseGO(geneList = gene_stat_clean,
                ont = "ALL",
                keyType = "SYMBOL",
                minGSSize = 3,
                maxGSSize = 800,
                pvalueCutoff = 0.05,
                verbose = TRUE,
                OrgDb = org.Hs.eg.db,
                pAdjustMethod = "BH")

dotplot(gse_GO, showCategory = 6, split = ".sign") + facet_grid(. ~ .sign)  + theme(axis.text.y = element_text(size = 8))
```

GO GSEA analysis reveals that E1A suppresses ribosome biogenesis and rRNA processing, but activates extracellular matrix, neuron differentiation, cell projection morphogenesis, and blood vessel development.
This indicates a pronounced reprogramming of the cellular state toward tissue remodeling and differentiation programs.
The suppression of ribosome genesis can be explained as part of an antiproliferative program, since cancer cells require active ribosome biogenesis for rapid cell division.
The enhancement of ECM- and differentiation-related pathways may be a sign that E1A disrupts the tumor transcriptional program, shifting the cell from a growth state to a state of remodeling and partial "de-oncogenization."


```{r}
# GO BP (Biological Process)
gse_GO_BP <- gseGO(geneList = gene_stat_clean,
                   ont = "BP",
                   keyType = "SYMBOL",
                   minGSSize = 3,
                   maxGSSize = 800,
                   pvalueCutoff = 0.05,
                   verbose = FALSE,
                   OrgDb = org.Hs.eg.db,
                   pAdjustMethod = "BH")

dotplot(gse_GO_BP, showCategory = 6, split = ".sign") + facet_grid(. ~ .sign) + theme(axis.text.y = element_text(size = 8))

# GO MF (Molecular Function)
gse_GO_MF <- gseGO(geneList = gene_stat_clean,
                   ont = "MF",
                   keyType = "SYMBOL",
                   minGSSize = 3,
                   maxGSSize = 800,
                   pvalueCutoff = 0.05,
                   verbose = FALSE,
                   OrgDb = org.Hs.eg.db,
                   pAdjustMethod = "BH")

dotplot(gse_GO_MF, showCategory = 6, split = ".sign") + facet_grid(. ~ .sign) + theme(axis.text.y = element_text(size = 8))

# GO CC (Cellular Component)
gse_GO_CC <- gseGO(geneList = gene_stat_clean,
                   ont = "CC",
                   keyType = "SYMBOL",
                   minGSSize = 3,
                   maxGSSize = 800, 
                   pvalueCutoff = 0.05,
                   verbose = FALSE,
                   OrgDb = org.Hs.eg.db,
                   pAdjustMethod = "BH")

dotplot(gse_GO_CC, showCategory = 6, split = ".sign") + facet_grid(. ~ .sign) + theme(axis.text.y = element_text(size = 8))
```

```{r}
# SYMBOL → ENTREZ
mapping_kegg <- mapIds(org.Hs.eg.db,
                       keys = names(gene_stat),
                       column = "ENTREZID",
                       keytype = "SYMBOL",
                       multiVals = "first")

# geneList with ENTREZ ID
gene_list_entrez <- gene_stat
names(gene_list_entrez) <- mapping_kegg
gene_list_entrez <- na.omit(gene_list_entrez)

# KEGG GSEA
gsea_kegg <- gseKEGG(geneList = gene_list_entrez,
                     organism = "hsa",
                     pvalueCutoff = 0.05,
                     pAdjustMethod = "BH")

gsea_kegg <- setReadable(gsea_kegg, 'org.Hs.eg.db', keyType = "ENTREZID")

dotplot(gsea_kegg, showCategory = 6, split = ".sign") + facet_grid(. ~ .sign) + theme(axis.text.y = element_text(size = 8))
```

KEGG shows downregulation of proteasome, RNA degradation, ribosome, ribosome biogenesis, nuclear transport, and related pathways.
This typically indicates a decrease in the cell's overall level of housekeeping activity: less protein synthesis, less ribosomogenesis, weaker control of RNA degradation, and possibly less nucleocytoplasmic trafficking.
For HCT116, this can be interpreted as a weakening of the proliferative/growth mode and a restructuring of the cellular state.
If phagosome, ECM-receptor interaction, neuroactive ligand-receptor interaction, and other signaling pathways are activated, this most often indicates that the cell is entering a state of remodeling, stress, and altered microenvironmental contacts and intercellular communication.

```{r}
#Transcription factors
if (!require("tftargets")) {
    install.packages("devtools")
    devtools::install_github("slowkow/tftargets")
}

gmt_file <- "TRED_db.gmt"
file_conn <- file(gmt_file, "w")

for (tf_name in names(TRED)) {
  target_genes <- TRED[[tf_name]]
  gmt_line <- paste(
    tf_name,                  
    "TRED_target_genes",      
    paste(target_genes, collapse = "\t"),  
    sep = "\t"
  )
  writeLines(gmt_line, file_conn)
}
close(file_conn)

gmt_to_TERM2GENE <- function(gmt_file) {
  readLines(gmt_file) %>% 
    strsplit("\t") %>% 
    lapply(function(x) {
      term <- x[1]
      genes <- x[-c(1, 2)]  
      data.frame(term = term, gene = genes)
    }) %>% 
    bind_rows()
}

TERM2GENE <- gmt_to_TERM2GENE("TRED_db.gmt")
head(TERM2GENE)

# GSEA on transcription factors
gsea_TRED <- GSEA(geneList = gene_list_entrez,
                  pvalueCutoff = 0.05,
                  pAdjustMethod = "none",
                  TERM2GENE = TERM2GENE)

gsea_TRED <- setReadable(gsea_TRED, 'org.Hs.eg.db', keyType = "ENTREZID")

dotplot(gsea_TRED, showCategory = 8, split = ".sign") + facet_grid(. ~ .sign)
gseaplot2(gsea_TRED, geneSetID = 1:3)
```

GSEA of the TRED transcription factor database
demonstrated that E1A expression leads to the activation of a wide range of transcription factors (NES 1.34 to 1.89), while no significantly suppressed TFs were detected.
This is consistent with literature data: E1A primarily functions as a transcriptional activator through interaction with p300/CBP coactivators and regulation of the E2F pathway.
E1A expression activates a wide range of transcription factors, including SMAD4, FOS, ATF2, ETS2, and others, indicating a profound reorganization of the cellular regulatory network.
In HCT116, this likely reflects not a single linear pathway, but a complex shift between proliferation, stress response, and differentiation programs.


```{r}
# ctrl vs dox+
keep <- colData(dds)$cell_line == "HCT116" & colData(dds)$condition == "ctrl"
dds_HCT116 <- dds[, keep]

table(dds_HCT116$e1a_status)
```

```{r}
#  e1a_status only
design(dds_HCT116) <- ~ e1a_status

# DESeq2
dds_HCT116 <- DESeq(dds_HCT116)

# e1a_status_positive_vs_negative
res_E1A_HCT116 <- results(dds_HCT116, name = "e1a_status_positive_vs_negative")
summary(res_E1A_HCT116)
```

```{r}
# padj < 0.05 и |log2FC| > 1
up <- sum(res_E1A_HCT116$padj < 0.05 & res_E1A_HCT116$log2FoldChange > 1, na.rm = TRUE)
down <- sum(res_E1A_HCT116$padj < 0.05 & res_E1A_HCT116$log2FoldChange < -1, na.rm = TRUE)
cat("Up-regulated:", up, "\n")
cat("Down-regulated:", down, "\n")
cat("Total DEGs:", up + down, "\n")
```

The effect of E1A is significantly stronger in sensitive HCT116 cells (183 DEGs) compared to the full model (117 DEGs).
In HCT116, E1A not only activates but also represses many genes (34 downregulated), whereas in the full model, downregulation is minimal (8 downregulated).
This indicates that the resistant cell line and/or oxaliplatin modify the response to E1A, attenuating its transcriptional effects.


```{r}
# Top-30 по padj
top_genes_HCT116 <- head(order(res_E1A_HCT116$padj), 30)

# VST
vsd_HCT116 <- vst(dds_HCT116, blind = FALSE)
top_data_HCT116 <- assay(vsd_HCT116)[top_genes_HCT116, ]
colnames(top_data_HCT116) <- ifelse(colData(dds_HCT116)$e1a_status == "negative", 
                                    "ctrl", "E1A")

# Heatmap
pheatmap(top_data_HCT116,
         scale = "row",
         main = "Top 30 DEGs: E1A effect in HCT116",
         fontsize_row = 6,
         fontsize_col = 8,
         angle_col = 0,         
         treeheight_row = 50,
         treeheight_col = 50,
         border_color = NA)   
```



```{r}
EnhancedVolcano(res_E1A_HCT116,
    lab = rownames(res_E1A_HCT116),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'E1A effect in HCT116 (positive vs negative)',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.0,
    labSize = 4.0)
```

In both models, E1A consistently activates the early response and AP-1-related program: JUNB, EGR1, ARC, and BRD2 are well-suited for this role.
This means that, regardless of the background, E1A triggers rapid transcriptional rearrangements associated with stress, the immediate response, and the regulation of proliferation.

```{r}
# Список генов
genes_of_interest <- c("JUNB", "EGR1", "DSP-AS1")

for (gene in genes_of_interest) {
  ge <- plotCounts(dds_HCT116, gene, intgroup = "e1a_status", returnData = TRUE)
  
  # Меняем названия для легенды
  ge$condition <- ifelse(ge$e1a_status == "negative", "ctrl", "E1A")
  
  p <- ggplot(ge, aes(x = condition, y = count, fill = condition)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.8) +
    scale_fill_manual(values = c("ctrl" = "lightblue", "E1A" = "salmon")) +
    labs(title = paste("Expression of", gene),
         x = "", y = "Normalized Counts") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 12, face = "bold"),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  
  print(p)
}
```



```{r}
gene_stat_HCT116 <- res_E1A_HCT116$stat
names(gene_stat_HCT116) <- rownames(res_E1A_HCT116)
gene_stat_HCT116 <- na.omit(gene_stat_HCT116)
gene_stat_HCT116 <- sort(gene_stat_HCT116, decreasing = TRUE)

# SYMBOL → ENTREZ
mapping_HCT116 <- mapIds(org.Hs.eg.db,
                         keys = names(gene_stat_HCT116),
                         column = "ENTREZID",
                         keytype = "SYMBOL",
                         multiVals = "first")

gene_stat_clean_HCT116 <- gene_stat_HCT116[!is.na(mapping_HCT116)]

# GSEA GO ALL 
gse_GO_HCT116 <- gseGO(geneList = gene_stat_clean_HCT116,
                       ont = "ALL",
                       keyType = "SYMBOL",
                       minGSSize = 3,
                       maxGSSize = 800,
                       pvalueCutoff = 0.05,
                       verbose = TRUE,
                       OrgDb = org.Hs.eg.db,
                       pAdjustMethod = "BH")

dotplot(gse_GO_HCT116, showCategory = 6, split = ".sign") + 
    facet_grid(. ~ .sign, scales = "free") + 
    theme(axis.text.y = element_text(size = 10)) +
    ggtitle("GSEA GO ALL: E1A effect in HCT116 (ctrl vs E1A)")

# KEGG
# SYMBOL → ENTREZ
mapping_kegg_HCT116 <- mapIds(org.Hs.eg.db,
                              keys = names(gene_stat_HCT116),
                              column = "ENTREZID",
                              keytype = "SYMBOL",
                              multiVals = "first")

# geneList with ENTREZ ID
gene_list_entrez_HCT116 <- gene_stat_HCT116
names(gene_list_entrez_HCT116) <- mapping_kegg_HCT116
gene_list_entrez_HCT116 <- na.omit(gene_list_entrez_HCT116)

# KEGG GSEA
gsea_kegg_HCT116 <- gseKEGG(geneList = gene_list_entrez_HCT116,
                            organism = "hsa",
                            pvalueCutoff = 0.05,
                            pAdjustMethod = "BH")

dotplot(gsea_kegg_HCT116, showCategory = 6, split = ".sign") + 
    facet_grid(. ~ .sign) + 
    theme(axis.text.y = element_text(size = 10)) +
    ggtitle("KEGG: E1A effect in HCT116 (ctrl vs E1A)")
```

When analyzed on 24 samples (the full model, including the resistant line and oxaliplatin treatment), the same pathways (e.g., ribosome biogenesis, extracellular matrix) were also enriched, but had higher p-values (p.adjust ~1e-3 – 1e-5) compared to analysis of HCT116 alone (p.adjust ~5e-08).
This indicates that the effect of E1A on the transcriptome is attenuated in the presence of additional factors (oxaliplatin resistance, treatment).
In pure HCT116, the effect of E1A is most pronounced and statistically reliable.

```{r}
# Preparing geneList for HCT116 (sensitive, ctrl vs E1A)
gene_stat_TF <- res_E1A_HCT116$stat
names(gene_stat_TF) <- rownames(res_E1A_HCT116)
gene_stat_TF <- na.omit(gene_stat_TF)
gene_stat_TF <- sort(gene_stat_TF, decreasing = TRUE)

# SYMBOL → ENTREZ
mapping_TF <- mapIds(org.Hs.eg.db,
                     keys = names(gene_stat_TF),
                     column = "ENTREZID",
                     keytype = "SYMBOL",
                     multiVals = "first")

gene_list_entrez_TF <- gene_stat_TF[!is.na(mapping_TF)]
names(gene_list_entrez_TF) <- mapping_TF[!is.na(mapping_TF)]

gmt_file <- "TRED_db.gmt"
file_conn <- file(gmt_file, "w")

for (tf_name in names(TRED)) {
  target_genes <- TRED[[tf_name]]
  gmt_line <- paste(tf_name, "TRED_target_genes", paste(target_genes, collapse = "\t"), sep = "\t")
  writeLines(gmt_line, file_conn)
}
close(file_conn)


gmt_to_TERM2GENE <- function(gmt_file) {
  readLines(gmt_file) %>% 
    strsplit("\t") %>% 
    lapply(function(x) {
      term <- x[1]
      genes <- x[-c(1, 2)]  
      data.frame(term = term, gene = genes)
    }) %>% 
    bind_rows()
}
TERM2GENE <- gmt_to_TERM2GENE("TRED_db.gmt")

# GSEA 
gsea_TRED_HCT116 <- GSEA(geneList = gene_list_entrez_TF,
                         pvalueCutoff = 0.05,
                         pAdjustMethod = "none",
                         TERM2GENE = TERM2GENE)

gsea_TRED_HCT116 <- setReadable(gsea_TRED_HCT116, 'org.Hs.eg.db', keyType = "ENTREZID")

dotplot(gsea_TRED_HCT116, showCategory = 8, split = ".sign") + 
    facet_grid(. ~ .sign) +
    ggtitle("TRED: E1A effect in HCT116 (ctrl vs E1A)")

gseaplot2(gsea_TRED_HCT116, geneSetID = 1:3)
```



```{r}
# Leave only resistant HCT116_oxpl_R cells with control (without oxaliplatin)
keep_R <- colData(dds)$cell_line == "HCT116_oxpl_R" & colData(dds)$condition == "ctrl"
dds_R <- dds[, keep_R]

design(dds_R) <- ~ e1a_status
dds_R <- DESeq(dds_R)

res_E1A_R <- results(dds_R, name = "e1a_status_positive_vs_negative")
summary(res_E1A_R)

# up/down |log2FC| > 1
up_R <- sum(res_E1A_R$padj < 0.05 & res_E1A_R$log2FoldChange > 1, na.rm = TRUE)
down_R <- sum(res_E1A_R$padj < 0.05 & res_E1A_R$log2FoldChange < -1, na.rm = TRUE)
cat("Up-regulated:", up_R, "\n")
cat("Down-regulated:", down_R, "\n")
```


```{r}
top_genes_R <- head(order(res_E1A_R$padj), 30)

vsd_R <- vst(dds_R, blind = FALSE)
top_data_R <- assay(vsd_R)[top_genes_R, ]
colnames(top_data_R) <- ifelse(colData(dds_R)$e1a_status == "negative", 
                               "ctrl", "dox+")

pheatmap(top_data_R,
         scale = "row",
         main = "Top 30 DEGs: E1A effect in resistant cells (ctrl vs dox+)",
         fontsize_row = 6,
         fontsize_col = 8,
         angle_col = 0,
         treeheight_row = 50,
         treeheight_col = 50,
         border_color = NA)
```

```{r}
EnhancedVolcano(res_E1A_R,
    lab = rownames(res_E1A_R),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'E1A effect in resistant cells',
    subtitle = 'HCT116_oxpl_R (ctrl vs dox+)',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.0,
    labSize = 4.0)
```

In oxaliplatin-sensitive HCT116 cells, E1A elicits a broader transcriptional response, whereas in resistant HCT116 oxpl-R cells the response is more constrained. TACSTD2 remains mildly induced but to a lower extent in the resistant background, EPHA2 is already highly expressed at baseline in resistant cells, and H2AC18 shows a stronger induction upon doxycycline in the resistant state, suggesting that E1A acts on a pre-rewired chromatin and signaling landscape. Overall, the smaller number of responsive genes in resistant cells indicates reduced transcriptional plasticity to E1A

```{r}
keep_heatmap <- colData(dds)$condition == "ctrl" & 
                colData(dds)$cell_line %in% c("HCT116", "HCT116_oxpl_R")

counts_heatmap <- counts(dds)[, keep_heatmap]

coldata_heatmap <- colData(dds)[keep_heatmap, ]

dds_heatmap <- DESeqDataSetFromMatrix(
    countData = counts_heatmap,
    colData = coldata_heatmap,
    design = ~ e1a_status + cell_line   # two factors
)

# VST 
vsd_heatmap <- vst(dds_heatmap, blind = FALSE)

top_genes <- head(rownames(res_E1A_HCT116)[order(res_E1A_HCT116$padj)], 30)
top_data <- assay(vsd_heatmap)[top_genes, ]

colnames(top_data) <- paste0(colData(dds_heatmap)$cell_line, "_", 
                             ifelse(colData(dds_heatmap)$e1a_status == "negative", "ctrl", "dox+"))

pheatmap(top_data,
         scale = "row",
         main = "Top 30 DEGs: E1A effect in sensitive vs resistant cells",
         fontsize_row = 6,
         fontsize_col = 8,
         angle_col = 45,
         treeheight_row = 50,
         treeheight_col = 50,
         border_color = NA)
```
```{r}
dds_ctrl <- dds[, colData(dds)$condition == "ctrl"]

genes_of_interest <- c("TACSTD2", "EPHA2", "H2AC18")

for (gene in genes_of_interest) {
  ge <- plotCounts(dds_ctrl, gene, intgroup = c("cell_line", "e1a_status"), returnData = TRUE)
  
  ge$group <- paste0(ge$cell_line, "\n", ifelse(ge$e1a_status == "negative", "ctrl", "dox+"))
  
  p <- ggplot(ge, aes(x = group, y = count, fill = e1a_status)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.8) +
    scale_fill_manual(values = c("negative" = "lightblue", "positive" = "salmon")) +
    labs(title = paste("Expression of", gene),
         x = "", y = "Normalized Counts") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 10, angle = 0),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 12, face = "bold"),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  
  print(p)
}
```

TACSTD2, EPHA2 and H2AC18 were particularly informative because they reflect different biological layers of the E1A response. TACSTD2 suggests a modest shift in epithelial/cell-surface state, EPHA2 indicates a pre-existing signaling plasticity in oxaliplatin-resistant cells, and H2AC18 points to a strong chromatin-level remodeling response upon doxycycline induction. Together, their behavior suggests that E1A acts on a pre-adapted resistant background rather than on a uniform cellular state.

```{r}
# GSEA 
gene_stat_R <- res_E1A_R$stat
names(gene_stat_R) <- rownames(res_E1A_R)
gene_stat_R <- na.omit(gene_stat_R)
gene_stat_R <- sort(gene_stat_R, decreasing = TRUE)

# SYMBOL → ENTREZ (как для чувствительных)
mapping_R <- mapIds(org.Hs.eg.db,
                    keys = names(gene_stat_R),
                    column = "ENTREZID",
                    keytype = "SYMBOL",
                    multiVals = "first")

gene_stat_clean_R <- gene_stat_R[!is.na(mapping_R)]

# GSEA GO ALL 
gse_GO_R <- gseGO(geneList = gene_stat_clean_R,
                  ont = "ALL",
                  keyType = "SYMBOL",
                  minGSSize = 3,
                  maxGSSize = 800,
                  pvalueCutoff = 0.05,
                  verbose = TRUE,
                  OrgDb = org.Hs.eg.db,
                  pAdjustMethod = "BH")

dotplot(gse_GO_R, showCategory = 6, split = ".sign") + 
    facet_grid(. ~ .sign, scales = "free") + 
    theme(axis.text.y = element_text(size = 8)) +
    ggtitle("GSEA GO ALL: E1A effect in resistant cells")

# KEGG 
gene_list_entrez_R <- gene_stat_R
names(gene_list_entrez_R) <- mapping_R
gene_list_entrez_R <- na.omit(gene_list_entrez_R)

gsea_kegg_R <- gseKEGG(geneList = gene_list_entrez_R,
                       organism = "hsa",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH")

dotplot(gsea_kegg_R, showCategory = 6, split = ".sign") + 
    facet_grid(. ~ .sign) + 
    theme(axis.text.y = element_text(size = 8)) +
    ggtitle("KEGG: E1A effect in resistant cells")
```

In resistant cells, E1A retains its ability to suppress ribosomogenesis and rRNA processing with high significance (p.adjust = 1e-05 - 1e-03), but activation of extracellular matrix and differentiation pathways is virtually absent. Instead, weak activation of immune and inflammatory pathways is observed. This indicates a partial, but incomplete, retention of E1A's effect in resistant cells.

