---
title: "E1A-induced chemosensitization"
author: "Elena Kalita"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)

library(DESeq2)
library(tximport)
library(ggplot2)
library(pheatmap)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(stringr)
library(DOSE)
library(dplyr)
```

# Aim of analysis

**E1A-induced chemosensitization**

Main question:

> How does E1A expression alter the transcriptional response of XCT116 cells to oxaliplatin?

For this we need interaction contrast:

``` text
(oxpl_E1A - ctrl_E1A) - (oxpl - ctrl)
```

That is, we are not simply comparing `oxpl_E1A vs oxpl`, but we are checking whether the response to oxaliplatin has changed in the presence of E1A.

This RMD includes two analyses:

1.  All 24 samples: cell line-corrected model.
2.  Only sensitive HCT116: ctrl, E1A, oxpl, oxpl_E1A.

# import Salmon quant.sf

```{r salmon-files}
samples <- list.files("Salmon quant", full.names = TRUE)
files <- file.path(samples, "quant.sf")
names(files) <- basename(samples)
```

```{r tximport}
tx2gene <- read.csv(gzfile("data/gencode.v47.metadata.HGNC.gz"), sep = "\t", header = FALSE)

txi_gene <- tximport(files, type = "salmon", tx2gene = tx2gene, countsFromAbundance = "no")
txi_gene_RPK <- txi_gene$counts / txi_gene$length
txi_gene_TPM <- tximport(files, type = "salmon", tx2gene = tx2gene, countsFromAbundance = "lengthScaledTPM")

txi_counts <- data.frame(txi_gene$counts)
txi_TPM <- data.frame(txi_gene_TPM$counts)

write.table(txi_counts, file = "raw_counts_all_samples.txt", sep = "\t", row.names = TRUE, quote = FALSE)
write.csv2(txi_counts, file = "raw_counts_all_samples.csv", row.names = TRUE, quote = FALSE)
write.csv2(txi_TPM, file = "TPM_all_samples.csv", row.names = TRUE, quote = FALSE)
```

# Metadata

```{r metadata}
metadata <- data.frame(
  full_condition = factor(rep(c("ctrl", "E1A", "oxpl", "oxpl_E1A"), each = 3, times = 2)),
  condition = factor(rep(c("ctrl", "oxpl"), each = 6, times = 2)),
  e1a_status = factor(rep(c("negative", "positive"), each = 3, times = 4)),
  cell_line = factor(rep(c("HCT116", "HCT116_oxpl_R"), each = 12)),
  stat = factor(rep(c("ctrl", "ctrl_E1A", "oxpl", "oxpl_E1A",
                      "R_ctrl", "R_ctrl_E1A", "R_oxpl", "R_oxpl_E1A"), each = 3))
)

rownames(metadata) <- colnames(txi_gene$counts)

metadata$condition <- relevel(metadata$condition, ref = "ctrl")
metadata$e1a_status <- relevel(metadata$e1a_status, ref = "negative")
metadata$cell_line <- relevel(metadata$cell_line, ref = "HCT116")

metadata$sample_label <- paste0(metadata$stat, "_rep", rep(1:3, times = 8))
metadata
```

# 1. All 24 samples

The full factorial model is used here:

``` text
~ cell_line * e1a_status * condition
```

It takes into account the main effects of cell line, E1A, and oxaliplatin,\
as well as all two- and three-way interactions.\
The main coefficient e1a_statuspositive.conditionoxpl shows\
how E1A modifies the response to oxaliplatin in HCT116.

## DESeq2 model: 24 samples

```{r deseq-24}
dds_24 <- DESeqDataSetFromTximport(
  txi = txi_gene,
  colData = metadata,
  design = ~ cell_line * e1a_status * condition
)

dds_24 <- DESeq(dds_24)
dds_24 <- dds_24[rowSums(counts(dds_24)) > 0, ]

resultsNames(dds_24)
```

## Main contrast: E1A changes oxaliplatin response, 24 samples

```{r result-24}
res_24 <- results(dds_24, name = "e1a_statuspositive.conditionoxpl", alpha = 0.05)
summary(res_24)

res_24_df <- as.data.frame(res_24)
res_24_df$gene <- rownames(res_24_df)

res_24_clean <- res_24_df[!is.na(res_24_df$padj) & !is.na(res_24_df$log2FoldChange), ]
res_24_sig <- res_24_clean[res_24_clean$padj < 0.05 & abs(res_24_clean$log2FoldChange) > 1, ]
res_24_up <- res_24_sig[res_24_sig$log2FoldChange > 1, ]
res_24_down <- res_24_sig[res_24_sig$log2FoldChange < -1, ]

metrics_24 <- data.frame(
  analysis = "24_samples",
  contrast = "E1A_changes_oxaliplatin_response",
  formula = "~ cell_line * e1a_status * condition",
  n_total_nonzero = nrow(res_24_df),
  n_with_pvalue = sum(!is.na(res_24_df$pvalue)),
  n_with_padj = sum(!is.na(res_24_df$padj)),
  n_sig_padj_0_05_absLFC_1 = nrow(res_24_sig),
  n_up = nrow(res_24_up),
  n_down = nrow(res_24_down)
)

metrics_24
```

```{r save-result-24}
write.csv(res_24_df, "E1A_chemosens_24_all_DESeq2_results.csv", row.names = FALSE)
write.csv(res_24_sig, "E1A_chemosens_24_significant_DEG.csv", row.names = FALSE)
write.csv(metrics_24, "E1A_chemosens_24_metrics.csv", row.names = FALSE)
```

## Top DEG: 24 samples

```{r top-deg-24}
top30_24 <- head(res_24_clean[order(res_24_clean$padj), ], 30)
top30_24[, c("gene", "log2FoldChange", "pvalue", "padj")]

write.csv(top30_24, "E1A_chemosens_24_top30_DEG.csv", row.names = FALSE)
```

## MA-plot: 24 samples

```{r ma-plot-24}
plotMA(res_24, main = "24 samples: E1A changes oxaliplatin response")
```

## VST and heatmap top-30: 24 samples

```{r heatmap-24, fig.width=11, fig.height=8}
vsd_24 <- vst(dds_24, blind = FALSE)

mat_24 <- assay(vsd_24)[top30_24$gene, ]
colnames(mat_24) <- metadata$sample_label

annotation_24 <- as.data.frame(colData(dds_24)[, c("cell_line", "condition", "e1a_status", "full_condition", "stat")])
rownames(annotation_24) <- metadata$sample_label

pheatmap(
  mat_24,
  scale = "row",
  annotation_col = annotation_24,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  fontsize_row = 8,
  fontsize_col = 7,
  main = "Top 30 DEG: E1A changes oxaliplatin response, 24 samples"
)
```

## Volcano plot: 24 samples

```{r volcano-24, fig.width=9, fig.height=8}
EnhancedVolcano(
  res_24,
  lab = rownames(res_24),
  x = "log2FoldChange",
  y = "pvalue",
  pCutoffCol = "padj",
  title = "Full DESeq model: E1A changes oxaliplatin response",
  FCcutoff = 1,
  pCutoff = 0.05,
  pointSize = 2.5,
  labSize = 4
)
```

## Boxplots for up- and down-regulated genes: 24 samples

We take the top up- and down-regulated genes from the DESeq2 result for the main interaction contrast. Here, `up` denotes positive `log2FoldChange`, `down` denotes negative `log2FoldChange`.

```{r boxplot-genes-24}
boxplot_up_24 <- head(res_24_up[order(res_24_up$padj), ], 6)
boxplot_down_24 <- head(res_24_down[order(res_24_down$padj), ], 6)

boxplot_genes_24 <- c(boxplot_up_24$gene, boxplot_down_24$gene)
boxplot_direction_24 <- c(rep("up", nrow(boxplot_up_24)), rep("down", nrow(boxplot_down_24)))

boxplot_table_24 <- data.frame(
  gene = boxplot_genes_24,
  direction = boxplot_direction_24
)

boxplot_table_24

write.csv(boxplot_table_24, "E1A_chemosens_24_boxplot_genes.csv", row.names = FALSE)
```

```{r boxplots-24, fig.width=10, fig.height=5}
dir.create("boxplots_E1A_chemosens_24", showWarnings = FALSE)

for (i in seq_along(boxplot_genes_24)) {
  gene <- boxplot_genes_24[i]
  direction <- boxplot_direction_24[i]
  
  ge <- plotCounts(dds_24, gene = gene, intgroup = "stat", returnData = TRUE)
  
  p <- ggplot(ge, aes(x = stat, y = count, fill = stat)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.8) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = paste("24 samples:", direction, "gene", gene),
      x = "",
      y = "Normalized counts"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  
  print(p)
  
  safe_gene <- gsub("[^A-Za-z0-9_.-]", "_", gene)
  ggsave(
    filename = file.path("boxplots_E1A_chemosens_24", paste0(direction, "_", safe_gene, "_boxplot.png")),
    plot = p,
    width = 10,
    height = 5,
    dpi = 300
  )
}
```

## GSEA ranking: 24 samples

For GSEA we use Wald test statistics (`stat`).

```{r gsea-ranking-24}
gene_stat_24 <- res_24$stat
names(gene_stat_24) <- rownames(res_24)
gene_stat_24 <- na.omit(gene_stat_24)
gene_stat_24 <- sort(gene_stat_24, decreasing = TRUE)

head(gene_stat_24)
```

## GSEA GO ALL: 24 samples

```{r gsea-go-all-24, fig.width=11, fig.height=8}
gse_GO_24_ALL <- gseGO(
  geneList = gene_stat_24,
  ont = "ALL",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(
  as.data.frame(gse_GO_24_ALL),
  "E1A_chemosens_24_GSEA_GO_ALL.csv",
  row.names = FALSE
)

dotplot(gse_GO_24_ALL, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO ALL, 24 samples") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO BP: 24 samples

```{r gsea-go-bp-24, fig.width=11, fig.height=8}
gse_GO_24_BP <- gseGO(
  geneList = gene_stat_24,
  ont = "BP",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_24_BP), "E1A_chemosens_24_GSEA_GO_BP.csv", row.names = FALSE)

dotplot(gse_GO_24_BP, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO BP, 24 samples") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO CC: 24 samples

```{r gsea-go-cc-24, fig.width=11, fig.height=7}
gse_GO_24_CC <- gseGO(
  geneList = gene_stat_24,
  ont = "CC",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_24_CC), "E1A_chemosens_24_GSEA_GO_CC.csv", row.names = FALSE)

dotplot(gse_GO_24_CC, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO CC, 24 samples") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO MF: 24 samples

```{r gsea-go-mf-24, fig.width=11, fig.height=7}
gse_GO_24_MF <- gseGO(
  geneList = gene_stat_24,
  ont = "MF",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_24_MF), "E1A_chemosens_24_GSEA_GO_MF.csv", row.names = FALSE)

dotplot(gse_GO_24_MF, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO MF, 24 samples") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA KEGG: 24 samples

```{r gsea-kegg-24, fig.width=10, fig.height=6}
mapping_24 <- mapIds(
  org.Hs.eg.db,
  keys = names(gene_stat_24),
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"
)

gene_list_entrez_24 <- gene_stat_24
names(gene_list_entrez_24) <- mapping_24

gene_list_entrez_24 <- gene_list_entrez_24[!is.na(names(gene_list_entrez_24))]
gene_list_entrez_24 <- gene_list_entrez_24[!duplicated(names(gene_list_entrez_24))]
gene_list_entrez_24 <- sort(gene_list_entrez_24, decreasing = TRUE)

gsea_kegg_24 <- gseKEGG(
  geneList = gene_list_entrez_24,
  organism = "hsa",
  keyType = "ncbi-geneid",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

gsea_kegg_24 <- setReadable(gsea_kegg_24, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

write.csv(as.data.frame(gsea_kegg_24), "E1A_chemosens_24_GSEA_KEGG.csv", row.names = FALSE)

dotplot(gsea_kegg_24, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA KEGG 24 samples") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

# 2. Only sensitive HCT116 cells

Now we leave only 12 HCT116 samples:

``` text
ctrl, E1A, oxpl, oxpl_E1A
```

Model:

``` text
~ e1a_status * condition
```

Main contrast:

``` text
(oxpl_E1A - ctrl_E1A) - (oxpl - ctrl)
```

## Subset HCT116

```{r subset-hct116}
metadata_HCT116 <- metadata[metadata$cell_line == "HCT116", ]
metadata_HCT116 <- droplevels(metadata_HCT116)

samples_HCT116 <- rownames(metadata_HCT116)

txi_HCT116 <- txi_gene
txi_HCT116$counts <- txi_gene$counts[, samples_HCT116]
txi_HCT116$abundance <- txi_gene$abundance[, samples_HCT116]
txi_HCT116$length <- txi_gene$length[, samples_HCT116]

metadata_HCT116$condition <- relevel(metadata_HCT116$condition, ref = "ctrl")
metadata_HCT116$e1a_status <- relevel(metadata_HCT116$e1a_status, ref = "negative")
metadata_HCT116$sample_label <- paste0(metadata_HCT116$stat, "_rep", rep(1:3, times = 4))

metadata_HCT116
```

## DESeq2 model: HCT116 only

```{r deseq-hct116}
dds_HCT116 <- DESeqDataSetFromTximport(
  txi = txi_HCT116,
  colData = metadata_HCT116,
  design = ~ e1a_status * condition
)

dds_HCT116 <- DESeq(dds_HCT116)
dds_HCT116 <- dds_HCT116[rowSums(counts(dds_HCT116)) > 0, ]

resultsNames(dds_HCT116)
```

## Main contrast: E1A changes oxaliplatin response, HCT116 only

```{r result-hct116}
res_HCT116 <- results(dds_HCT116, name = "e1a_statuspositive.conditionoxpl", alpha = 0.05)
summary(res_HCT116)

res_HCT116_df <- as.data.frame(res_HCT116)
res_HCT116_df$gene <- rownames(res_HCT116_df)

res_HCT116_clean <- res_HCT116_df[!is.na(res_HCT116_df$padj) & !is.na(res_HCT116_df$log2FoldChange), ]
res_HCT116_sig <- res_HCT116_clean[res_HCT116_clean$padj < 0.05 & abs(res_HCT116_clean$log2FoldChange) > 1, ]
res_HCT116_up <- res_HCT116_sig[res_HCT116_sig$log2FoldChange > 1, ]
res_HCT116_down <- res_HCT116_sig[res_HCT116_sig$log2FoldChange < -1, ]

metrics_HCT116 <- data.frame(
  analysis = "HCT116_only",
  contrast = "E1A_changes_oxaliplatin_response",
  formula = "~ e1a_status * condition",
  n_total_nonzero = nrow(res_HCT116_df),
  n_with_pvalue = sum(!is.na(res_HCT116_df$pvalue)),
  n_with_padj = sum(!is.na(res_HCT116_df$padj)),
  n_sig_padj_0_05_absLFC_1 = nrow(res_HCT116_sig),
  n_up = nrow(res_HCT116_up),
  n_down = nrow(res_HCT116_down)
)

metrics_HCT116
```

```{r save-result-hct116}
write.csv(res_HCT116_df, "E1A_chemosens_HCT116_all_DESeq2_results.csv", row.names = FALSE)
write.csv(res_HCT116_sig, "E1A_chemosens_HCT116_significant_DEG.csv", row.names = FALSE)
write.csv(metrics_HCT116, "E1A_chemosens_HCT116_metrics.csv", row.names = FALSE)
```

## Top DEG: HCT116 only

```{r top-deg-hct116}
top30_HCT116 <- head(res_HCT116_clean[order(res_HCT116_clean$padj), ], 30)
top30_HCT116[, c("gene", "log2FoldChange", "pvalue", "padj")]

write.csv(top30_HCT116, "E1A_chemosens_HCT116_top30_DEG.csv", row.names = FALSE)
```

## MA-plot: HCT116 only

```{r ma-plot-hct116}
plotMA(res_HCT116, main = "HCT116: E1A changes oxaliplatin response")
```

## VST and heatmap top-30: HCT116 only

```{r heatmap-hct116, fig.width=9, fig.height=8}
vsd_HCT116 <- vst(dds_HCT116, blind = FALSE)

mat_HCT116 <- assay(vsd_HCT116)[top30_HCT116$gene, ]
colnames(mat_HCT116) <- metadata_HCT116$sample_label

annotation_HCT116 <- as.data.frame(colData(dds_HCT116)[, c("condition", "e1a_status", "full_condition", "stat")])
rownames(annotation_HCT116) <- metadata_HCT116$sample_label

pheatmap(
  mat_HCT116,
  scale = "row",
  annotation_col = annotation_HCT116,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  fontsize_row = 8,
  fontsize_col = 8,
  main = "Top 30 DEG: E1A changes oxaliplatin response, HCT116 sensitive only"
)
```

## Volcano plot: HCT116 only

```{r volcano-hct116, fig.width=9, fig.height=8}
EnhancedVolcano(
  res_HCT116,
  lab = rownames(res_HCT116),
  x = "log2FoldChange",
  y = "pvalue",
  pCutoffCol = "padj",
  title = "E1A-induced chemosensitization",
  FCcutoff = 1,
  pCutoff = 0.05,
  pointSize = 2.5,
  labSize = 4
)
```

## Boxplots for up- and down-regulated genes: HCT116 only

We take the top up- and down-regulated genes from the DESeq2 result for interaction contrast only in sensitive HCT116.

```{r boxplot-genes-hct116}
boxplot_up_HCT116 <- res_HCT116_up[order(res_HCT116_up$padj), ]
boxplot_down_HCT116 <- res_HCT116_down[order(res_HCT116_down$padj), ]

boxplot_genes_HCT116 <- c(boxplot_up_HCT116$gene, boxplot_down_HCT116$gene)
boxplot_direction_HCT116 <- c(
  rep("up", nrow(boxplot_up_HCT116)),
  rep("down", nrow(boxplot_down_HCT116))
)

boxplot_table_HCT116 <- data.frame(
  gene = boxplot_genes_HCT116,
  direction = boxplot_direction_HCT116
)

boxplot_table_HCT116
write.csv(boxplot_table_HCT116, "E1A_chemosens_HCT116_boxplot_genes.csv", row.names = FALSE)
```

```{r boxplots-hct116, fig.width=8, fig.height=5}
dir.create("boxplots_E1A_chemosens_HCT116", showWarnings = FALSE)

for (i in seq_along(boxplot_genes_HCT116)) {
  gene <- boxplot_genes_HCT116[i]
  direction <- boxplot_direction_HCT116[i]
  
  ge <- plotCounts(dds_HCT116, gene = gene, intgroup = "stat", returnData = TRUE)
  
  p <- ggplot(ge, aes(x = stat, y = count, fill = stat)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.8) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = paste("HCT116:", direction, "gene", gene),
      x = "",
      y = "Normalized Counts"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  
  print(p)
  
  safe_gene <- gsub("[^A-Za-z0-9_.-]", "_", gene)
  ggsave(
    filename = file.path("boxplots_E1A_chemosens_HCT116", paste0(direction, "_", safe_gene, "_boxplot.png")),
    plot = p,
    width = 8,
    height = 5,
    dpi = 300
  )
}
```

## GSEA ranking: HCT116 only

```{r gsea-ranking-hct116}
gene_stat_HCT116 <- res_HCT116$stat
names(gene_stat_HCT116) <- rownames(res_HCT116)
gene_stat_HCT116 <- na.omit(gene_stat_HCT116)
gene_stat_HCT116 <- sort(gene_stat_HCT116, decreasing = TRUE)

head(gene_stat_HCT116)
```

## GSEA GO ALL: HCT116 only

```{r gsea-go-all-hct116, fig.width=11, fig.height=9}
gse_GO_HCT116_ALL <- gseGO(
  geneList = gene_stat_HCT116,
  ont = "ALL",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_HCT116_ALL), "E1A_chemosens_HCT116_GSEA_GO_ALL.csv", row.names = FALSE)

dotplot(gse_GO_HCT116_ALL, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO ALL: HCT116 sensitive only") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO BP: HCT116 only

```{r gsea-go-bp-hct116, fig.width=11, fig.height=9}
gse_GO_HCT116_BP <- gseGO(
  geneList = gene_stat_HCT116,
  ont = "BP",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_HCT116_BP), "E1A_chemosens_HCT116_GSEA_GO_BP.csv", row.names = FALSE)

dotplot(gse_GO_HCT116_BP, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO BP: HCT116 sensitive only") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO CC: HCT116 only

```{r gsea-go-cc-hct116, fig.width=10, fig.height=6}
gse_GO_HCT116_CC <- gseGO(
  geneList = gene_stat_HCT116,
  ont = "CC",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_HCT116_CC), "E1A_chemosens_HCT116_GSEA_GO_CC.csv", row.names = FALSE)

dotplot(gse_GO_HCT116_CC, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO CC: HCT116 sensitive only") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA GO MF: HCT116 only

```{r gsea-go-mf-hct116, fig.width=10, fig.height=7}
gse_GO_HCT116_MF <- gseGO(
  geneList = gene_stat_HCT116,
  ont = "MF",
  keyType = "SYMBOL",
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = org.Hs.eg.db,
  pAdjustMethod = "BH"
)

write.csv(as.data.frame(gse_GO_HCT116_MF), "E1A_chemosens_HCT116_GSEA_GO_MF.csv", row.names = FALSE)

dotplot(gse_GO_HCT116_MF, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO MF: HCT116 sensitive only") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## GSEA KEGG: HCT116 only

```{r gsea-kegg-hct116, fig.width=10, fig.height=6}
mapping_HCT116 <- mapIds(
  org.Hs.eg.db,
  keys = names(gene_stat_HCT116),
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"
)

gene_list_entrez_HCT116 <- gene_stat_HCT116
names(gene_list_entrez_HCT116) <- mapping_HCT116

gene_list_entrez_HCT116 <- gene_list_entrez_HCT116[!is.na(names(gene_list_entrez_HCT116))]
gene_list_entrez_HCT116 <- gene_list_entrez_HCT116[!duplicated(names(gene_list_entrez_HCT116))]
gene_list_entrez_HCT116 <- sort(gene_list_entrez_HCT116, decreasing = TRUE)

gsea_kegg_HCT116 <- gseKEGG(
  geneList = gene_list_entrez_HCT116,
  organism = "hsa",
  keyType = "ncbi-geneid",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

gsea_kegg_HCT116 <- setReadable(gsea_kegg_HCT116, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

write.csv(as.data.frame(gsea_kegg_HCT116), "E1A_chemosens_HCT116_GSEA_KEGG.csv", row.names = FALSE)

dotplot(gsea_kegg_HCT116, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign) +
  labs(title = "GSEA GO KEGG: HCT116 sensitive only") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

# General table of metrics

```{r final-metrics}
metrics_all <- rbind(metrics_24, metrics_HCT116)
metrics_all

write.csv(metrics_all, "E1A_chemosens_all_metrics.csv", row.names = FALSE)
```

# 
